---
title: "Presentation code"
author: "Maciej Dobrzyński"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
    code_folding: show
---

# Intro

This notebook accompanies *MIC training: Modern data analysis in R/RStudio*. The corresponding presentation is [here](../../presentation-main.html).

To convert this notebook into HTML (to *knit* it), type `make` in the command line of this folder or click **Knit** icon in the top bar of the RStudio editor.


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
               cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# data.table

Define a `data.table`:

```{r, echo = T}
library(data.table)
dtBabies = data.table(name= c("Jackson", "Emma", "Liam", "Ava"), 
                      gender = c("M", "F", "M", "F"), 
                      year2011= c(74.69, NA, 88.24, 81.77), 
                      year2012=c(84.99, NA, NA, 96.45), 
                      year2013=c(91.73, 75.74, 101.83, NA),
                      year2014=c(95.32, 82.49, 108.23, NA),
                      year2015=c(107.12, 93.73, 119.01, 105.65))
dtBabies
```
Select specific records/rows:

```{r, echo = T}
dtBabies[gender == 'M']
```

Select specific variables/columns:

```{r, echo = T}
dtBabies[, .(name, gender, year2015)]
```
Calculate the mean of a single column:

```{r, echo = T}
dtBabies[, 
         .(meanWeight = mean(year2015))]
```

Calculate the mean of a single column by gender:

```{r, echo = T}
dtBabies[, 
         .(meanWeight = mean(year2015)), 
         by = gender]
```

# data.table - reference by name

Define a list of string variables with column names:

```{r, echo = T}
lCol = list(meas = 'weight',
            time = 'year',
            group = c('name', 'gender'),
            timeLast = 'year2015')
lCol
```

Select records with `gender` column equal to `M`.

The name of the gender column is stored in one of the list elements.

```{r, echo = T}
lCol$group
```

We need the second entry of that list element.

```{r, ehco = T}
lCol$group[2]
```

To let `data.table` know that we want to provide column names via pre-defined string variables, we need to use function `get` in the `i` position.

Hence, to select rows:

```{r, echo = T}
dtBabies[get(lCol$group[[2]]) == 'M']
```

To select specific columns we can provide them this way:

```{r, echo = T}
myColumns = c(lCol$group[[1]], lCol$timeLast)
dtBabies[, myColumns, with = F]
```
Or simpler (note the double dot):

```{r, echo = T}
dtBabies[, ..myColumns]
```

To aggregate by calculating the mean:

```{r, echo = T}
dtBabies[, 
         .(meanWeight = mean(get(lCol$timeLast))), 
         by = c(lCol$group[2])]
```

# Wide to long format
Our data table is in the wide format. To convert it to long format, use function `melt` from the `data.table` package.

Provide the names of **identification** (`id.vars`) and **measure** variables (`measure.vars`). If none are provided, `melt` guesses them automatically, which may result in a wrong conversion. Both variables can be given as strings with column names, or as column numbers.

The original data frame contains missing values; `na.rm=T` omits them in the long-format table.

```{r, echo = T}
dtBabiesLong = data.table::melt(dtBabies, 
                                id.vars = c('name', 'gender'), 
                                measure.vars = 3:7,
                                variable.name = 'year', 
                                value.name = 'weight',
                                na.rm = T)
head(dtBabiesLong, n = 5L)
```

# Long to wide

The function `dcast` converts from long to wide format. The function has a so-called *formula interface* that specifies a combination of variables that uniquely identify a row.

Note that because some combinations of `name + gender + year` do not exist, the `dcast` function will introduce `NAs`.

```{r, echo = T}
dtBabiesWide = data.table::dcast(dtBabiesLong, 
                                 name + gender ~ year, 
                                 value.var = 'weight')

dtBabiesWide
```

One way to use the formula interface with string variables is to create the formula string explicitly by concatenating individual string pieces:

```{r, echo = T}
as.formula(
  paste0(
    lCol$group[[1]], "+", 
    lCol$group[[2]], "~", 
    lCol$time))
```

```{r, echo = T}
dtBabiesWide2 = data.table::dcast(dtBabiesLong, 
                                  as.formula(
                                    paste0(
                                      lCol$group[[1]], "+", 
                                      lCol$group[[2]], "~", 
                                      lCol$time)), 
                                  value.var = 'weight')

all.equal(dtBabiesWide,
          dtBabiesWide2)
```

# data.table - IO

Fast reading of the files; use `nThread` option to take advantage of multiple threads and read files even faster! 

Compressed files can be read directly without decompressing them first!

```{r, echo = T}
myDT = data.table::fread(file = "../../practicals/data/m1_allGF_wExpDescr.csv.gz", 
                         nThread = 2)
```

Write as a CSV file and then compress it:

```{r, echo = T}
# Write a subset
myFilePath = "testOutput.csv"
data.table::fwrite(x = myDT[Stim_Treat == "EGF" & Metadata_T < 10],
                   file = myFilePath, 
                   row.names = F)

R.utils::gzip(myFilePath, overwrite = T)
```

```{r, echo = F}
# Clean
rm(myDT,
   myFilePath,
   dtBabiesWide,
   dtBabiesWide2)
```

# Plotting with ggplot2

Let's plot weight of individual babies over time.

```{r, echo = F}
library(ggplot2)

ggplot2::ggplot(dtBabiesLong, 
                aes(x = year, 
                    y = weight)) +
  geom_line()
```


Oops, the plotting function doesn't know how to link the points. The logical way to link them is by the `name` column. We need to tell `ggplot2` how to group the data.

Also, to avoid hard-coding column names we use `aes_string` instead of `aes`. 

The data will be plotted as lines, with additional dots to indicate data points.

```{r, echo = T}
ggplot2::ggplot(dtBabiesLong, 
                aes_string(x = lCol$time, 
                           y = lCol$meas, 
                           group = lCol$group[1])) +
  geom_line() +
  geom_point()
```

In order to produce facets per `gender`, we use function `facet_wrap`. 

```{r, echo = T}
ggplot2::ggplot(dtBabiesLong, 
                aes_string(x = lCol$time, 
                           y = lCol$meas, 
                           group = lCol$group[1])) +
  geom_line() +
  geom_point() +
  facet_wrap(lCol$group[2])
```
## Summaries

```{r, echo = T}
ggplot2::ggplot(dtBabiesLong, 
                     aes_string(x = lCol$time, 
                                y = lCol$meas, 
                                group = lCol$group[1])) +
  geom_line() +
  geom_point() +
  facet_wrap(lCol$group[2]) +
  stat_summary(fun = mean, 
               aes(group=1), 
               geom = "line", 
               colour = 'red')
```

## Trends

```{r, echo = T}
ggplot2::ggplot(dtBabiesLong, 
                aes_string(x = lCol$time, 
                           y = lCol$meas)) +
  geom_point() +
  facet_wrap(lCol$group[2]) +
  geom_smooth(method = "lm",
              formula = y ~ x,
              colour = "red")
```

Oops, regression analysis requires numerical variables! Here, the time in the `year` column is a string and is treated as a categorical variable.

Extract numerical value from the `year` column and assign it to a `yearNum` column. The result of the string substituting function `gsub` is also a string, hence we need to convert the result to a number using `as.numeric` function.

```{r, echo = T}
dtBabiesLong[,
             yearNum := as.numeric(gsub("year", "", get(lCol$time)))]
```

```{r, echo = T}
ggplot2::ggplot(dtBabiesLong, 
                aes_string(x = "yearNum", 
                           y = lCol$meas)) +
  geom_point() +
  facet_wrap(lCol$group[2]) +
  geom_smooth(method = "lm",
              formula = y ~ x,
              colour = "red")
```

Shaded regions correspond to 95% CI.

## Customisations

Lot's of different themes and customisations, e.g. `theme_bw()`, `theme_minimal()`, etc.

Also [check](https://mran.microsoft.com/snapshot/2017-02-04/web/packages/ggthemes/vignettes/ggthemes.html) the `ggthemes` package for additional themes and colour scales.

```{r, echo = T}
p1 = ggplot2::ggplot(dtBabiesLong, 
                     aes_string(x = lCol$time, 
                                y = lCol$meas, 
                                group = lCol$group[1])) +
  geom_line() +
  geom_point() +
  stat_summary(fun = mean, 
               aes(group=1), 
               geom = "line", 
               colour = 'red') +
  facet_wrap(lCol$group[2]) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, 
                                   hjust = 1))
```

Note, that the plot was assigned to a variable `p1` and there is no graphics output. This is useful if we want to do something additional with the plot. For example, we can add additional `ggplot` layers or save the plot to a file.

To display the plot simply invoke the variable:

```{r, echo = T}
p1
```


## Interactive plots

Now when we have the plot stored in `p1` variable, we can convert it to an interactive plot using `plotly` package.

```{r, echo = T}
library(plotly)
plotly::ggplotly(p1)
```

```{r, echo = F}
# Clean
rm(p1)
```


# Functions

Let's test the `calcStats()` [function](../calcStats.R) on our dataset.

```{r, echo = T}
source("../calcStats.R")

calcStats(inDt = dtBabiesLong, 
          inMeasVar = "weight",
          inGroupName = "gender",
          inRobust = T)
```

## Unit testing

```{r, echo = T}
library(testthat)

# calculated result
resCalc = data.table(meas = 1:9)
resTrue = data.table(meanMeas = 5.0)

# Test should pass; NO message will be produced
testthat::expect_equal(calcStats(inDt = resCalc,
                                 inMeasVar = "meas"), 
                       resTrue)
```

```{r, echo = T, eval = F}
# Test should not pass; error message will appear
resTrue = data.table(meanMeas = 4.0)

testthat::expect_equal(calcStats(inDt = resCalc,
                                 inMeasVar = "meas"), 
                       resTrue)
```

## Unit testing with test folder

Create a `test` folder and create individual files with tests. The name of these files should start with `test_` prefix, e.g. `test_calcStats.R`.

```{r, echo = T}
library(testthat)

testthat::test_dir("../tests")
```

```{r, echo = F}
# Celan
rm(dtBabies,
   dtBabiesLong,
   lCol,
   resCalc,
   resTrue,
   myColumns,
   calcStats)
```

## Profiling

[Source code](../ex_profiling_time_01.R) with examples for `profvis`.

[Source code](../ex_profiling_time_02.R) with examples for `microbenchmark`.

## Debugging

[Source code](../ex_debug_browser.R) with examples.

# Vectorization

Suppose we need to add 10M numbers to each other. This would be a *classical* way of doing it using a for loop:

```{r, echo = T}
nNum = 1e7
vA = rnorm(nNum)
vB = rnorm(nNum)
vCloop = rep(0., nNum)

system.time(
  for(ii in 1:nNum) {
    vCloop[ii] = vA[ii] + vB[ii]
  }  
)
```

Instead, R provides a convenient operator `+` that also works with vectors!

```{r, echo = T}
system.time({vCvec = vA + vB})
```

The result of two operations is exactly the same, but the latter is way faster!

```{r, echo = T}
all.equal(vCloop, vCvec)
```

```{r, echo=FALSE}
# Clean
rm(nNum,
   vA, vB, vCloop, vCvec)
```

# For loops

For loops can be fast **if** you follow some rules:

1. Don’t use a loop when a vectorized alternative exists (see the previous slide).
2. Don't increase the size of your objects, e.g. using `cbind`, `rbind`, during the loop! Instead, pre-allocate memory by predefining an object to hold the result. 

Example of an inefficient loop: at every iteration of the loop, `myRes` increases in size because a new element is added to this vector.

```{r, echo = T}
nNum = 1e4
vA = rnorm(nNum)
vB = rnorm(nNum)

myRes = c()
system.time(
  for(ii in 1:nNum) {
    myRes = c(myRes, vA[ii] + vB[ii])
  }  
)
```

Example of an efficient loop: the result vector `vC` is pre-allocated and has the same size throughout the iteration of the loop. The only thing that changes is the content of that vector!

```{r}
# Pre-allocated vector where the result will be written to
vC = rep(0., nNum)

system.time(
  for(ii in seq_along(vA)) {
    vC[ii] = vA[ii] + vB[ii]
  }  
)
```

```{r}
all.equal(myRes, vC)
```

```{r, echo=FALSE}
rm(nNum,
   vA, vB, myRes, vC)
```

The result is the same, but pre-allocating the `vC` vector speeds up the computation by two orders of magnitude!

The difference between the computation time increases with the size of the vector because without pre-allocating the memory, R is copying the existing vector before increasing its size with the `c` operation.

# Apply functions

[R notebook](../ex_apply/ex_apply.html) with examples.

# Parallel computations using foreach

[R notebook](../ex_parallel_foreach/ex_parallel_foreach.html) with example calculations.

# Command-line parameters

[Source code](../ex_optparse.R) with examples.