---
title: "Parallel computations with foreach"
author: "Maciej Dobrzyński"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
    code_folding: hide
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Intro

To convert this notebook into HTML (to *knit* it), type `make` in the command line of this folder or click **Knit** icon in the top bar of the RStudio editor.

The example CPU intensive calculation taken from [here](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html)

In this example a random forest model is calculated for a predictor matrix of nRow x nCol random gaussian numbers and a response vector with two classes.

```{r, echo=T}
## Load required packages
require(randomForest)
require(foreach)
require(doParallel)

## Define variables
# Number of rows and columns for a predictor matrix
nRow = 1000
nCol = 5

# Create a predictor matrix with normally distributed random numbers
mX <- matrix(runif(nRow * nCol), nRow)

# Create a vector with numbers that belong to two factors, 1 & 2
# These are classes for machine learning classification, i.e. a response vector
vY <- gl(2, as.integer(round(nRow/2)))

# Create a vector with iterator;
# The foreach will loop over that vector 4 times and
# will create a tree ensemble of 4000 decision trees.
vIter = rep(4000, 4)
```

# Sequential calculation

The `randomForest` function will be repeated 4 times and the result will be assigned to `resRF` variable which holds random forests from each iteration combined into a single tree ensemble using `randomForest::combine` function.

```{r, echo = T}
system.time({
  resRF = foreach(ntree=vIter,
                  .combine = randomForest::combine) %do%
    randomForest(mX, vY, 
                 ntree=ntree)
})
```

The resulting tree ensemble.

```{r, echo = T}
resRF
```

# Parallel with 2 cores

We need to let each independent process know to include the `randomForest` package.

To specify the number of cores use `registerDoParallel()` function from `doParallel` package. Here, we will use 2 cores.

```{r, echo = T}
numCores = 2
doParallel::registerDoParallel(numCores)
```

To check the number of available cores on the system use `parallel::detectCores()`.

```{r, echo = T}
parallel::detectCores()
```

Note the `stopImplicitCluster()` to de-register the cluster. Provides a way for clean exit.

```{r, echo = T}
system.time({
  resRF = foreach(ntree=vIter, 
                  .combine = randomForest::combine,
                  .packages='randomForest') %dopar%
    randomForest(mX, vY, 
                 ntree=ntree)
})

doParallel::stopImplicitCluster()
```

# Parallel with 4 cores

```{r, echo = T}
numCores = 4
doParallel::registerDoParallel(numCores)

system.time({
  resRF = foreach::foreach(ntree=vIter,
                           .combine = randomForest::combine,
                           .packages='randomForest') %dopar%
    randomForest::randomForest(mX, vY, 
                               ntree=ntree)
})
doParallel::stopImplicitCluster()
```

# Conclusion

The reason the elapsed time does not increase linearly with the number of processes is the following. 

First, The calculation of tree ensembles takes place in parallel, however, the subsequent `combine` is a single-process operation. This is an important part of any parallel computation. Gathering data from all processes also takes time and it might be a significant part of the execution time! 

Second, modern CPUs have cores with [HyperThreading](https://en.wikipedia.org/wiki/Hyper-threading), which presents a single physical core of the CPU as two logical cores to the system. It allows to perform several instructions simultaneously on a single physical core. While it can improve the performance of some calculations, pushing a lot of the same type of calculations at the same time might actually hamper the performance. 