---
title: "Reading files from a folder"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    highlight: kate
    code_folding: hide
---


```{r setup, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

# Intro

To convert this notebook into HTML (to *knit* it), type `make` in the command line of this folder or click **Knit** icon in the top bar of the RStudio editor.

Demonstrate reading files from a folder and combining them into a single object

```{r, echo = T}
library(data.table)

# List all CSV files in the folder; store them in a character vector
vFiles = list.files(path = "../ex_apply_data/.", 
                    pattern = "csv", 
                    full.names = T)
```

# Combine into a single object

```{r, echo = T}
lMyExp = lapply(vFiles, fread)
dtMyExp = rbindlist(lMyExp)

head(dtMyExp)
```

```{r, echo = T}
tail(dtMyExp)
```

**Problem**: combined `data.table` contains data from all experiments one after another. Which one is which???

# Combine with experiment info

```{r, echo = T}
lMyExp = lapply(vFiles, function(inFile) {
  
  # Read a single file
  locDT = fread(inFile)
  
  # Add a column with the file name
  locDT[,
        fileName := inFile]
})

dtMyExp = rbindlist(lMyExp)

head(dtMyExp)
```

**Improvement**: extract experiment number from the `fileName` column using `gsub` function and [regular expressions](https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf).

The file name is of the form `xx/xx/exp_01.csv`. We would like to extract the number after `exp_` string. The regular expression captures a group with at least one digit (`[0-9]+`).

```{r, echo = T}
dtMyExp[,
        expName := as.integer(gsub(".*exp_([0-9]+).*", "\\1", fileName))]

# remove the unnecessary column fileName
dtMyExp[,
        fileName := NULL]

head(dtMyExp)
```

```{r, echo = T}
tail(dtMyExp)

```

